The role of cognateness in native spoken word recognition

### 2025-02-07 manuscript/manuscript.tex
Minor styling changes

Agency (Project PID2021- 123416NB-I00 financed by MCIN/ AEI/
10.13039/501100011033 / FEDER, UE) and the Economic and Social Research
Council (ESRC) (ES/S010947/1, UK). GGC was supported by a FPI research
contract (PRE2019-088165). We are grateful to Irina Lepadatu and Nicola
Gillen, and to the Speech Acquisition and Perception research group for
their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, Pamela
Miller, and Cristina Cuadrado for their technical support. We also thank
all families and infants who participated in the experiments. This study
was conducted according to guidelines laid down in the Declaration of
Helsinki, and was approved by the Drug Research Ethical Committee (CEIm)
of the IMIM Parc de Salut Mar, reference 2020/9080/I. The data and code
necessary to reproduce the analyses presented here are publicly
accessible, as are the materials necessary to attempt to replicate the
findings. The analyses presented here were not preregistered. Data,
code, and materials for this research are available at the following
URLs: https://osf.io/hy984/,
https://github.com/gongcastro/translation-elicitation.}}

@@

When some German speakers listened to the song \emph{The Power} by SNAP!
(\emph{Coyote Ugly}, 2000), many of them misheard the line ``I've got
the power'' as \emph{Agatha Bauer}. When Spanish speakers listen to the
line ``Circumvent your thick ego'' from the song \emph{Pictures} by
System of a Down (\emph{Steal this Album!}, 2002), they tend to hear
\emph{Sácame de aquí, King-Kong} {[}take me out of here, King-Kong{]}.
Outrageous as these examples might sound, the reader may know a few
cases in their own native language. This auditory illusion is common
across languages and cultures and can feel quite real, often inevitable
(\citeproc{ref-dembeck2015oberflachenubersetzung}{Dembeck, 2015};
\citeproc{ref-efimova2018homophonic}{Efimova et al., 2018}). In
Japanese, this phenomenon takes the name \emph{Soramimi} (lit. ``empty
ear''). From a linguistic point of view, Soramimi is a particular case
of homophonic translation: words or phrases that appear in the speech
stream in one language are translated into similar-sounding words and
phrases in a different language, without necessarily preserving the
meaning (\citeproc{ref-gasparov2006semen}{Gasparov, 2006}). Homophonic
translation has mostly received attention as a perceptual curiosity, and
as a literary figure, intentionally employed by translators to preserve
the meter and ``sound structure'' (e.g., rhymes) of an original text in
the language of translation (e.g.,
\citeproc{ref-levick2016translating}{Levick, 2016};
\citeproc{ref-pilshchikov2016semiotics}{Pilshchikov, 2016})\footnote{The
  rationale behind this translation approach is to ``sacrifice{[}s{]}
  fidelity to the image rather than the melodiousness of verse''
  (\citeproc{ref-briusov1911ot}{Briusov, 1911}). For instance, Edith
  Grossman's translation of \emph{El Quijote}
  (\citeproc{ref-cervantes}{Cervantes \& Grossman, 2002}) present a
  similar case, in which formal aspects of the original text in Spanish
  are favoured to some degree over equivalence to English, even
  sometimes keeping Spanish words from the original text in the English
  translation (e.g., \emph{Señor}, \emph{ínsula}).}. Little attention
has been paid to homophonic translation as a psycholinguistic event.

In one of the few available studies concerning homophonic translation as
a psycholinguistic phenomenon, Otake
(\citeproc{ref-otake2007interlingual}{2007}) analysed 194 instances of
Soramimi, broadcasted between 1992 and 2007 by the Japanese TV show
\emph{Soramimi hour}, in which examples of Soramimi were presented to
the audience for comedy purposes. These instances consisted of
homophonic translations of English song lyrics to words and phrases in
Japanese. The vast majority of the reported instances were phrases
(96\%) and few were single words (4\%). There was relatively high
variability in the degree to which phonetic features of the presented
English lyrics were preserved, with some resulting Japanese words or
phrases sharing high overlap with the original English strings of
sounds, and some sharing very little overlap. An analysis of the few
instances of homophonic translations of single words revealed three main
phonological processes that could explain how Japanese listeners
reconstructed the English input to generate the Japanese words:
insertion (e.g., \emph{cry} /\textipa{kraI}/ to \emph{kurai},
/\textipa{kuRai}/ {[}dark{]}), deletion (e.g., \emph{go},
/\textipa{goU}/ to \emph{go}, /go/ {[}go{]}, and alternation (e.g.,
\emph{low}, /\textipa{loU}/ to \emph{rou}, /\textipa{Roo}/ {[}wax{]}).
This suggests that the reported homophonic translations can be
explained, to some extent, as the result of the Japanese listeners
accommodating the strings of English sounds to the Japanese phoneme
inventory and Japanese phonotactics
(\citeproc{ref-dupoux1999epenthetic}{Dupoux et al., 1999};
\citeproc{ref-peperkamp2008perceptual}{Peperkamp et al., 2008}). The
phonological processes described by Otake
(\citeproc{ref-otake2007interlingual}{2007}) do not take place in a
vaccum, but rather in the context of a lexicon. As words or phrases in
the native lexicon are activated by their acoustic similarity with the
non-native speech stream, they trigger a series of mechanisms that
result in the selection of one of the candidate word or phrase in the
native language. In the present work, we investigated the interplay
between phonological similarity and one of the main mechanisms involved
in the dynamics of lexical selection: phonological neighbourhood
density. The phonological neighbourhood density of a given word-form
refers to the number of words-forms that differ in a single phoneme from
the former. For instance, the English word \emph{cat} hbelongs to a
dense phonological neighbourhood, with 50 English phonological
neighbours (e.g., hat, bat, cap), while the word \emph{charger} belongs
to a sparse phonological neighbourhood, with six neighbours (e.g.,
charged,larger, charter, charmer). Phonological neighbourhood density
affects lexical processing. For instance, words from dense
neighbourhoods are recognized more slowly and less accuractely in the
auditory modality than words from sparse neighbourhoods
(\citeproc{ref-dufour2010phonological}{Dufour \& Frauenfelder, 2010};
e.g., \citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998};
\citeproc{ref-vitevitch2006clustering}{M. Vitevitch, 2006};
\citeproc{ref-vitevitch1998words}{Vitevitch \& Luce, 1998}). In this
study, we examined how listeners may activate lexical representations in
their native language when listening to an unfamiliar language.

Although homophonic translations do not necessarily take place with the
preservation of meaning---which results in outrageous mistranslations
like the aforementioned ones---homophonic translations sometimes may
lead to correct translations. For example, imagine an English native
listening to Dutch for the first time. This listener may encounter the
word /\textipa{"pINgUIn}/ (\emph{pinguin}), which sounds similar to its
English translation /\textipa{"pENgwIn}/ (penguin). Given enough
phonological similarity between the non-native word and its translation
in the native language, a naïve listener may succeed at word
comprehension when listening to an unfamiliar language. Such
phonological similarity between translation equivalents---known as
\emph{cognateness}---is common across many languages, and is often due
to typological closeness and/or socio-historical events involving the
speakers of these languages (e.g., migration, social contact). For
example, Romance languages such as Spanish and Catalan share many
form-similar translation equivalents
(\citeproc{ref-schepens2012distributions}{Schepens et al., 2012}), as in
the case of \emph{puerta} and \emph{porta} (\emph{door} in Spanish and
Catalan, respectively). Given no prior knowledge of either language, a
Spanish native speaker is likely to be much more successful at correctly
translating Catalan \emph{porta} than English \emph{door}, due to the
phonological and (orthographic similarity) of the former to the
equivalent in their native language.

@@

We further investigated the effect of phonological competitors. Even if
the presented word and its target translation share high phonological
similarity, participants may still provide incorrect translations if the
presented word also shares high phonological similarity with other words
in the native language. Words with denser phonological neighbourhoods
are recognised more slowly and less accurately than words with sparser
phonological neighbourhoods (\citeproc{ref-dufour2003lexical}{Dufour \&
Peereman, 2003}; \citeproc{ref-goldinger1989priming}{Goldinger et al.,
1989}; \citeproc{ref-hamburger1996phonological}{Hamburger \& Slowiaczek,
1996}; \citeproc{ref-luce1990similarity}{Luce et al., 1990};
\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}). This is
especially true if such neighbours share higher phonological similarity
with the presented word, or are lexically more frequent than the target
word (\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}). Such
interference effect of phonological neighbourhood density has also been
reported across languages (\citeproc{ref-weber2004lexical}{Weber \&
Cutler, 2004}). We hypothesised that the size of the facilitatory role
of phonological similarity would be inversely proportional to the amount
of higher-frequency phonological neighbours of the presented word in the
target language. In the case of non-native speech recognition, the
presence of cross-linguistic pairs which are phonologically similar but
differ in meaning (e.g., false friends) may act as distractors during
lexical access, obstructing the selection of the appropriate target
translation in the listener's lexicon. For instance, Otwinowska \&
Szewczyk (\citeproc{ref-otwinowska2019more}{2019}) reported that false
friends were disadvantaged relative to non-cognates by Polish second
language learners, in contrast to cognates which were known better. It
is therefore important to investigate the joint effect of both cognates
and false friends when investigating the effect cross-linguistic
phonological similarity has on word recognition in a foreign language.
This could shed light on available strategies and challenges associated
with the early stages of second language acquisition. For instance, one
could expect English participants to incorrectly translate the Spanish
word \emph{botón} as \emph{bottom} instead of as its correct translation
\emph{button}, due to combined effect of the close phonological
similarity between \emph{botón} and \emph{bottom} along with the high
lexical frequency of \emph{bottom} relative to \emph{button}. To test
this prediction, we developed a lexical frequency-dependent measure of
cross-language phonological neighbourhood density, in which a neighbour
is counted only if it is higher frequency and is one phoneme apart from
the presented word. If the phonological neighbourhood density of the
target translation affects participants' performance negatively, this
would suggest that competitors in the native language affect recognition
of non-native words during recognition of foreign speech. We conducted a
series of three experiments to test these predictions.

@@

We collected data from 71 British English-native adults living in United
Kingdom (\emph{Mean} = 21.76 years, \emph{SD} = 2.15, Range = 18-26, 46
female). Data collection took place from June 04th, 2020 to June 25th,
2020. Participants were recruited via Prolific (5£ compensation) and
SONA (compensation in academic credits). Participants gave informed
consent before providing any data. The study was conducted in accordance
with ethical standards of the Declaration of Helsinki and the protocol
was by the University of Oxford Medical Sciences Inter-Divisional
Research Ethics Committee (IDREC) (R60939/RE007). Participants were
asked to complete the experiment using a laptop in a quiet place with
good internet connection.

@@

\toprule
\multicolumn{1}{l}{} &  & \multicolumn{2}{c}{Age} &  \\ 
\cmidrule(lr){3-4}
\multicolumn{1}{l}{} & N\textsuperscript{\textit{1}} & Mean ± SD  & Range & L2 \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 1} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & 35 (8) & $21.80$ ± $2.08$ & $18$–$26$ & Russian (1) \\ 
cat-ENG & 36 (4) & $21.72$ ± $2.25$ & $18$–$25$ & French (1), German (1), Italian (1), Punjabi (1), Several (1) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 2} \\ 
\midrule\addlinespace[2.5pt]
cat-SPA & 33 (12) & $21.85$ ± $3.00$ & $18$–$33$ & French (9), German (1), Italian (2) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 3} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & 32 (1) & $21.72$ ± $2.59$ & $18$–$26$ & German (2), Japanese (1) \\ 
cat-ENG & 32 (0) & $22.31$ ± $2.39$ & $18$–$26$ & Cantonese (1), Irish (1) \\ 
\bottomrule

@@

We created two lists of input words to be presented to participants in
the auditory modality: one in Catalan and one in Spanish. Words in the
Catalan list were 6.67 phonemes long on average (\emph{SD} = 2.06, Range
= 2-11), and the orthographic forms of their English translations (which
participants had to type) were 5.12 characters long on average
(\emph{SD} = 1.56, Range = 3-9). Words in the Spanish list were 7.27
phonemes long on average (\emph{SD} = 2.05, Range = 3-13), and the
orthographic form of their English translations were 5.29 characters
long on average (\emph{SD} = 1.77, Range = 3-12).

@@

For each word pair in the Catalan and Spanish lists, we defined three
predictors of interest: the lexical frequency of the correct English
translation (\emph{Frequency}), the phonological similarity between the
presented word and their correct English translation
(\emph{Similarity}), and the presented word's number of cross-language
phonological neighbours (\emph{CLPN}) in English. \emph{Frequency} was
included as a nuance predictor, under the hypothesis that---keeping
other predictors constant---participants would provide correct responses
more often when the correct translations are high frequency words in
their native language. Lexical frequencies of correct translations were
extracted from SUBTLEX-UK (\citeproc{ref-van2014subtlex}{Van Heuven et
al., 2014}), and transformed to Zipf scores. Translations without a
lexical frequency value were excluded from data analysis (2 in the
Catalan list, 5 in the Spanish list). \emph{Similarity}, our main
predictor of interest, was calculated as the Levenshtein similarity
between the X-SAMPA transcriptions of each pair of translations using
the \texttt{stringdist} R package (\citeproc{ref-van2014stringdist}{van
der Loo, 2014}). The Levenshtein distance computes the edit distance
between two character strings---in this case, two phonological
transcriptions---by counting the number of additions, deletions, and
substitutions necessary to make both strings identical
(\citeproc{ref-levenshtein1966binary}{Levenshtein, 1966}). We divided
this edit distance by the number of characters in the longest X-SAMPA
transcription of the translation pair. This transformation accounts for
the fact that Levenshtein distance tends to increase with the length of
the transcriptions. For interpretability, we subtracted this proportion
from one, so that values closer to one correspond to higher similarity
between phonological transcriptions, and values closer to zero
correspond to lower similarity between phonological transcriptions. For
example, the \emph{table} (\texttt{teIb@l})-\emph{mesa} (\texttt{mesa})
translation pair had 17\% similarity, while the \emph{train}
(\texttt{trEIn})-\emph{tren} (\texttt{tREn}) translation pair had 60\%
similarity. We calculated the number of \emph{CLPN} for each Catalan and
Spanish word by counting the number of English words with same or higher
lexical frequency, and whose phonological transcription (in X-SAMPA
format) differed in up to one phoneme from that of the presented Catalan
or Spanish word. Lexical frequencies and phonological transcriptions
were extracted from the multilingual database CLEARPOND
(\citeproc{ref-marian2012clearpond}{Marian et al., 2012})\footnote{{[}Phonological
  transcriptions in CLEARPOND were generated from eSPEAK,
  \url{http://espeak.sourceforge.net/}{]}}. Table~\ref{tbl-stimuli}
summarises the lexical frequency, phonological neighbourhood density and
phonological overlap of the words included in the Catalan and the
Spanish lists.

@@

The experiment was implemented online using Psychopy/Pavlovia
(\citeproc{ref-peirce2019psychopy2}{Peirce et al., 2019}). Participants
accessed the study from a link provided by Prolific or SONA and
completed the experiment from an internet browser (Chrome or Mozilla).
After giving their consent for participating, participants answered a
series of questions about their demographic status, their language
background, and the set up they were using for completing the study.
Then, participants completed the experimental task. Participants were
informed that they would listen to a series of pre-recorded words in
either Catalan or Spanish. They were instructed to listen to each word,
guess its meaning in English, and type their answer as soon as possible.
Participants were randomly assigned to the Catalan or Spanish lists.
Participants in the Catalan list were presented with 83 trials, and
participants in the Spanish list were presented with 99 trials.

Each trial started with a yellow fixation point presented during one
second on the centre of the screen over a black background. After one
second, the audio started playing while the dot remained being displayed
until the audio ended. Upon the offset of the fixation point and audio,
participants were prompted to write their answer by a
\texttt{\textgreater{}} symbol. Typed letters were displayed on the
screen in real time to provide visual feed-back to participants.
Participants were allowed to correct their answer. Then, participants
pressed the \texttt{RETURN/ENTER} key to confirm their answer and start
a new trial.

\begin{figure}[H]

{\centering \pandocbounded{\includegraphics[keepaspectratio]{_assets/img/design.png}}

@@

\caption{Schematic representation of a trial in the experimental task.
The trial stars with the presentation of a fixation point in the centre
of the screen (yellow dot). After 1,000 ms, the auditory word was
presented while the fixation point remained on screen. After the offset
of the audio, the fixation point disappeared and a visual prompt
(\texttt{\textgreater{}}) was presented. Participants then wrote their
answer and clicked \texttt{RETURN}, making the end of the trial.}

@@

We collected data for a total of 6,446 trials completed by 72 unique
participants. We excluded trials in which participants did not enter any
text (\emph{n} = 72), in which a response in a language other than
English was provided (e.g., ``\texttt{agua}'', \emph{n} = 51), in which
participants did not provide a whole word (e.g., ``\texttt{f}'',
\emph{n} = 5), and in which participants added comments to the
experimenter (e.g., ``\texttt{unsure}'', \emph{n} = 13). In addition, we
excluded data from participants that self-rated their oral and/or
written skills in Catalan and Spanish, or any other second language as
four or higher in a five-point scale (\emph{n} = 2), were diagnosed with
a language (\emph{n} = 1), or did not contribute more than 80\% of valid
trials (\emph{n} = 9).

>>>>

Agency (Project PGC2018-101831-B-I00 financed by Spanish Ministry of
Science Innovation and Universities (MCIU), Spanish Research Agency
(AEI) and the European Regional Development Fund (ERDF), and by the
Economic and Social Research Council (ESRC) (ES/S010947/1, UK). GGC was
supported by a FPI research contract (PRE2019-088165). SS was supported
by XXXXXXXXXXXXX. KP was supported by XXXXXXXXXXXXXXXXX. NSG was
supported by an ICREA Academia 2019 award from the Catalan Institution
for Research and Advanced Studies (ICREA). We are grateful to Irina
Lepadatu and Nicola Gillen, and to the Speech Acquisition and Perception
research group for their helpful feedback. We thank Xavier Mayoral,
Silvia Blanch, Pamela Miller, and Cristina Cuadrado for their technical
support. We also thank all families and infants who participated in the
experiments. This study was conducted according to guidelines laid down
in the Declaration of Helsinki, and was approved by the Drug Research
Ethical Committee (CEIm) of the IMIM Parc de Salut Mar, reference
2020/9080/I. The data and code necessary to reproduce the analyses
presented here are publicly accessible, as are the materials necessary
to attempt to replicate the findings. The analyses presented here were
not preregistered. Data, code, and materials for this research are
available at the following URLs: \url{https://osf.io/hy984/},
\url{https://github.com/gongcastro/translation-elicitation}.}}

@@

When some German speakers listen to the song \emph{The Power} by SNAP!
(\emph{Coyote Ugly}, 2000), many of them mishear the line ``I've got the
power'' as \emph{Agatha Bauer}. When Spanish speakers listen to the line
``Circumvent your thick ego'' from the song \emph{Pictures} by System of
a Down (\emph{Steal this Album!}, 2002), they tend to hear \emph{Sácame
de aquí, King-Kong} {[}take me out of here, King-Kong{]}. Outrageous as
these examples might sound, the reader may know a few cases in their own
native language. This auditory illusion is common across languages and
cultures and can feel quite real, often inevitable
(\citeproc{ref-dembeck2015oberflachenubersetzung}{Dembeck, 2015};
\citeproc{ref-efimova2018homophonic}{Efimova et al., 2018}). In
Japanese, this phenomenon takes the name \emph{Soramimi} (lit. ``empty
ear''). From a linguistic point of view, Soramimi is a particular case
of homophonic translation: words or phrases that appear in the speech
stream in one language are translated into similar-sounding words and
phrases in a different language, without necessarily preserving the
meaning (\citeproc{ref-gasparov2006semen}{Gasparov, 2006}). Homophonic
translation has mostly received attention as a perceptual curiosity, and
as a literary figure, intentionally employed by translators to preserve
the meter and ``sound structure'' (e.g., rhymes) of an original text in
the language of translation (e.g.,
\citeproc{ref-levick2016translating}{Levick, 2016};
\citeproc{ref-pilshchikov2016semiotics}{Pilshchikov, 2016})\footnote{The
  rationale behind this translation approach is to ``sacrifice{[}s{]}
  fidelity to the image rather than the melodiousness of verse''
  (\citeproc{ref-briusov1911ot}{Briusov, 1911}). For instance, Edith
  Grossman's translation of \emph{El Quijote}
  (\citeproc{ref-cervantes}{Cervantes \& Grossman, 2002}) present a
  similar case, in which formal aspects of the original text in Spanish
  are favoured to some degree over equivalence to English, even
  sometimes keeping Spanish words from the original text in the English
  translation (e.g., \emph{Señor}, \emph{ínsula}).}. Little attention
has been paid to homophonic translation from a language processing
perspective.

One exception is the study by Otake
(\citeproc{ref-otake2007interlingual}{2007}), who analysed 194 instances
of Soramimi, broadcasted between 1992 and 2007 by the Japanese TV show
\emph{Soramimi hour}, in which examples of Soramimi were presented to
the audience for comedy purposes. These instances consisted of
homophonic translations of English song lyrics to words and phrases in
Japanese. The vast majority of the reported instances were phrases
(96\%) and a few, single words (4\%). There was relatively high
variability in the degree to which phonetic features of the presented
English lyrics were preserved, with some resulting Japanese words or
phrases sharing high overlap with the original English strings of
sounds, and some sharing very little overlap. An analysis of the few
instances of homophonic translations of single words revealed three main
phonological processes that could explain how Japanese listeners
reconstructed the English input to generate the Japanese words:
insertion (e.g., \emph{cry} /\textipa{kraI}/ to \emph{kurai}
/\textipa{kuRai}/ {[}dark{]}), deletion (e.g., \emph{go} /\textipa{goU}/
to \emph{go} /go/ {[}\emph{Go}, a board game{]}, and alternation (e.g.,
\emph{low} /\textipa{loU}/ to \emph{rou} /\textipa{Roo}/ {[}wax{]}).
This suggests that the reported homophonic translations can be
explained, to some extent, as the result of the Japanese listeners
accommodating the strings of English sounds to the Japanese phoneme
inventory and Japanese phonotactics
(\citeproc{ref-dupoux1999epenthetic}{Dupoux et al., 1999};
\citeproc{ref-peperkamp2008perceptual}{Peperkamp et al., 2008}).

The phonological processes described by Otake
(\citeproc{ref-otake2007interlingual}{2007}) do not take place in a
vaccum, but rather in the context of a lexicon. As words or phrases in
the native lexicon are activated by their acoustic similarity with the
non-native speech stream, they trigger a series of mechanisms that
result in the selection of one of the candidate word or phrase in the
native language. In the present work, we investigated the interplay
between phonological similarity and one of the main mechanisms involved
in the dynamics of lexical selection: phonological neighbourhood
density. The phonological neighbourhood density of a given word-form
refers to the number of words-forms that differ in a single phoneme from
the it. For instance, the English word \emph{cat} belongs to a dense
phonological neighbourhood, with 50 English phonological neighbours
(e.g., hat, bat, cap), while the word \emph{charger} belongs to a sparse
phonological neighbourhood, with six neighbours (e.g., charged, larger,
charter, charmer) (\citeproc{ref-marian2012clearpond}{Marian et al.,
2012}). Phonological neighbourhood density affects lexical processing.
For instance, words from dense neighbourhoods are recognized more slowly
and less accuractely in the auditory modality than words from sparse
neighbourhoods (\citeproc{ref-luce1998recognizing}{Luce \& Pisoni,
1998}; \citeproc{ref-vitevitch1998words}{Vitevitch \& Luce, 1998}; e.g.,
\citeproc{ref-dufour2010phonological}{Dufour \& Frauenfelder, 2010};
\citeproc{ref-vitevitch2006clustering}{M. Vitevitch, 2006}). In this
study, we examined how listeners may activate lexical representations in
their native language when listening to an unfamiliar language.

Although homophonic translations rarely involve preservation of meaning,
they sometimes lead to correct translations. For example, imagine an
English native listening to Dutch for the first time. This listener may
encounter the word /\textipa{"pINgUIn}/ (\emph{pinguin}), which sounds
similar to its English translation /\textipa{"pENgwIn}/ (penguin). Given
enough phonological similarity between the non-native word and its
translation in the native language, a naïve listener may succeed at word
comprehension when listening to an unfamiliar language. Such
phonological similarity between translation equivalents---known as
\emph{cognateness}---is common across many languages, and is often due
to typological closeness and/or socio-historical events involving the
speakers of these languages (e.g., migration, social contact). For
example, Romance languages such as Spanish and Catalan share many
form-similar translation equivalents
(\citeproc{ref-schepens2012distributions}{Schepens et al., 2012}), as in
the case of \emph{puerta} and \emph{porta} (\emph{door} in Spanish and
Catalan, respectively). Given no prior knowledge of either language, a
Spanish native speaker is likely to be much more successful at correctly
translating Catalan \emph{porta} than English \emph{door}, due to the
phonological and (orthographic similarity) of the former to the
equivalent in their native language.

@@

We further investigated the effect of phonological competition. Even if
the presented word and its target translation share high phonological
similarity, participants may still provide incorrect translations if the
presented word also shares high phonological similarity with other words
in the native language. Words with denser phonological neighbourhoods
are recognised more slowly and less accurately than words with sparser
phonological neighbourhoods (\citeproc{ref-dufour2003lexical}{Dufour \&
Peereman, 2003}; \citeproc{ref-goldinger1989priming}{Goldinger et al.,
1989}; \citeproc{ref-hamburger1996phonological}{Hamburger \& Slowiaczek,
1996}; \citeproc{ref-luce1990similarity}{Luce et al., 1990};
\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}). This is
especially true if such neighbours share higher phonological similarity
with the presented word, or are lexically more frequent than the target
word (\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}). Such
interference effect of phonological neighbourhood density has also been
reported across languages (\citeproc{ref-weber2004lexical}{Weber \&
Cutler, 2004}). We hypothesised that the size of the facilitatory role
of phonological similarity would be inversely proportional to the amount
of higher-frequency phonological neighbours of the presented word in the
target language. In the case of non-native speech recognition, the
presence of cross-linguistic pairs which are phonologically similar but
differ in meaning (e.g., false friends) may act as distractors during
lexical access, obstructing the selection of the appropriate target
translation in the listener's lexicon. For instance, Otwinowska \&
Szewczyk (\citeproc{ref-otwinowska2019more}{2019}) reported that false
friends were disadvantaged relative to non-cognates by Polish second
language learners, in contrast to cognates which were known better. It
is therefore important to investigate the joint effect of both cognates
and false friends when investigating the effect that cross-linguistic
phonological similarity has on word recognition in a foreign language.
This could shed light on available strategies and challenges associated
with the early stages of second language acquisition. For instance, one
might expect English participants to incorrectly translate the Spanish
word \emph{botón} as \emph{bottom} instead of as its correct translation
\emph{button}, due to the combined effect of the close phonological
similarity between \emph{botón} and \emph{bottom} along with the high
lexical frequency of \emph{bottom} relative to \emph{button}. To test
this prediction, we developed a lexical frequency-dependent measure of
cross-language phonological neighbourhood density, in which a neighbour
is counted only if it is higher frequency and is one phoneme apart from
the presented word. If the phonological neighbourhood density of the
target translation affects participants' performance negatively, this
would suggest that competitors in the native language affect recognition
of non-native words during recognition of foreign speech. We conducted a
series of three experiments to test these predictions.

@@

We collected data from 71 British English-native adults living in the
United Kingdom (\emph{Mean} = 21.76 years, \emph{SD} = 2.15, Range =
18-26, 46 female). Data collection took place from June 4th, 2020 to
June 25th, 2020. Participants were recruited via Prolific (£5
compensation) and SONA (compensation in academic credits). Participants
gave informed consent before providing any data. The study was conducted
in accordance with ethical standards of the Declaration of Helsinki and
the protocol was by the University of Oxford Medical Sciences
Inter-Divisional Research Ethics Committee (IDREC) (R60939/RE007).
Participants were asked to complete the experiment using a laptop in a
quiet place with good internet connection.

@@

\toprule
\multicolumn{1}{l}{} &  & \multicolumn{2}{c}{Age} &  \\ 
\cmidrule(lr){3-4}
\multicolumn{1}{l}{} & N\textsuperscript{\textit{1}} & Mean ± SD  & Range & Second language \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 1} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & 35 (8) & $21.80$ ± $2.08$ & $18$–$26$ & Russian (1) \\ 
cat-ENG & 36 (4) & $21.72$ ± $2.25$ & $18$–$25$ & French (1), German (1), Italian (1), Punjabi (1), Several (1) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 2} \\ 
\midrule\addlinespace[2.5pt]
cat-SPA & 33 (12) & $21.85$ ± $3.00$ & $18$–$33$ & French (9), German (1), Italian (2) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 3} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & 32 (1) & $21.72$ ± $2.59$ & $18$–$26$ & German (2), Japanese (1) \\ 
cat-ENG & 32 (0) & $22.31$ ± $2.39$ & $18$–$26$ & Cantonese (1), Irish (1) \\ 
\bottomrule

@@

We created two lists of words to be presented to participants in the
auditory modality: one in Catalan and one in Spanish. Words in the
Catalan list were 6.67 phonemes long on average (\emph{SD} = 2.06, Range
= 2-11), and the orthographic forms of their English translations (which
participants had to type) were 5.12 characters long on average
(\emph{SD} = 1.56, Range = 3-9). Words in the Spanish list were 7.27
phonemes long on average (\emph{SD} = 2.05, Range = 3-13), and the
orthographic form of their English translations were 5.29 characters
long on average (\emph{SD} = 1.77, Range = 3-12).

@@

For each word pair in the Catalan and Spanish lists, we defined three
predictors of interest: the lexical frequency of the correct English
translation (\emph{Frequency}), the phonological similarity between the
presented word and its correct English translation (\emph{Similarity}),
and the presented word's number of cross-language phonological
neighbours (\emph{CLPN}) in English. \emph{Frequency} was included as a
nuisance predictor, under the hypothesis that---keeping other predictors
constant---participants would provide correct responses more often when
the correct translations are high frequency words in their native
language. Lexical frequencies of correct translations were extracted
from SUBTLEX-UK (\citeproc{ref-van2014subtlex}{Van Heuven et al.,
2014}), and transformed to Zipf scores. Translations without a lexical
frequency value were excluded from data analysis (2 in the Catalan list,
5 in the Spanish list). \emph{Similarity}, our main predictor of
interest, was calculated as the Levenshtein similarity between the
X-SAMPA transcriptions of each pair of translations using the
\texttt{stringdist} R package (\citeproc{ref-van2014stringdist}{van der
Loo, 2014}). The Levenshtein distance computes the edit distance between
two character strings---in this case, two phonological
transcriptions---by counting the number of additions, deletions, and
substitutions necessary to make both strings identical
(\citeproc{ref-levenshtein1966binary}{Levenshtein, 1966}). We divided
this edit distance by the number of characters in the longest X-SAMPA
transcription of the translation pair. This transformation accounts for
the fact that Levenshtein distance tends to increase with the length of
the transcriptions. For interpretability, we subtracted this proportion
from one, so that values closer to one correspond to higher similarity
between phonological transcriptions, and values closer to zero
correspond to lower similarity between phonological transcriptions. For
example, the \emph{table} (\texttt{teIb@l})-\emph{mesa} (\texttt{mesa})
translation pair had 17\% similarity, while the \emph{train}
(\texttt{trEIn})-\emph{tren} (\texttt{tREn}) translation pair had 60\%
similarity. We calculated the number of \emph{CLPN} for each Catalan and
Spanish word by counting the number of English words with same or higher
lexical frequency, and whose phonological transcription (in X-SAMPA
format) differed by up to one phoneme from that of the presented Catalan
or Spanish word. Lexical frequencies and phonological transcriptions
were extracted from the multilingual database CLEARPOND
(\citeproc{ref-marian2012clearpond}{Marian et al., 2012})\footnote{Phonological
  transcriptions in CLEARPOND were generated from eSPEAK,
  \url{http://espeak.sourceforge.net/}}{]}. Table~\ref{tbl-stimuli}
summarises the lexical frequency, phonological neighbourhood density and
phonological overlap of the words included in the Catalan and the
Spanish lists.

@@

The experiment was implemented online using Psychopy/Pavlovia
(\citeproc{ref-peirce2019psychopy2}{Peirce et al., 2019}). Participants
accessed the study from a link provided by Prolific or SONA and
completed the experiment from an internet browser (Chrome or Mozilla).
After giving their consent for participating, participants answered a
series of questions about their demographic status, their language
background, and the set up they were using for completing the study.
Then, participants completed the experimental task. Participants were
informed that they would listen to a series of pre-recorded words in
either Catalan or Spanish. They were instructed to listen to each word,
guess its meaning in English, and type their answer as soon as possible.
Participants were randomly assigned to the Catalan or Spanish lists.
Participants in the Catalan list were presented with 83 trials, and
participants in the Spanish list were presented with 99
trials\footnote{Stimuli lists were built from a pool of audio recordings
  made for a different study. The different number of trials in the
  Catalan and Spanish lists is due to more Spanish audios being
  available than Catalan audios.}.

Each trial started with a yellow fixation point presented during one
second on the centre of the screen over a black background. After one
second, the audio started playing while the dot remained being displayed
until the audio ended. Upon the offset of the fixation point and audio,
participants were prompted to write their answer by a
\texttt{\textgreater{}} symbol. Typed letters were displayed on the
screen in real time to provide visual feed-back to participants.
Participants were allowed to correct their answer. Then, participants
pressed the \texttt{RETURN/ENTER} key to confirm their answer and start
a new trial. Figure~\ref{fig-design} illustrates the time course of a
trial in the translation elicitation task.

\pandocbounded{\includegraphics[keepaspectratio]{_assets/img/design.png}}

@@

\caption{\label{fig-design}Schematic representation of a trial in the
experimental task. The trial stars with the presentation of a fixation
point in the centre of the screen (yellow dot). After 1,000 ms, the
auditory word was presented while the fixation point remained on screen.
After the offset of the audio, the fixation point disappeared and a
visual prompt (\texttt{\textgreater{}}) was presented. Participants then
wrote their answer and clicked \texttt{RETURN}, making the end of the
trial.}

@@

We collected data for a total of 6,446 trials completed by 72
participants. We excluded trials in which participants did not enter any
text (\emph{n} = 72), in which a response in a language other than
English was provided (e.g., ``\texttt{agua}'', \emph{n} = 51), in which
participants did not provide a whole word (e.g., ``\texttt{f}'',
\emph{n} = 5), and in which participants added comments to the
experimenter (e.g., ``\texttt{unsure}'', \emph{n} = 13). In addition, we
excluded data from participants that self-rated their oral and/or
written skills in Catalan and Spanish, or any other second language as
four or higher in a five-point scale (\emph{n} = 2), were diagnosed with
a language disorder (\emph{n} = 1), or did not contribute more than 80\%
of valid trials (\emph{n} = 9).
